---
title: "Compare biodiversity indicators from the b3gbi package"
author: "Emma Cartuyvels, Ward Langeraert, Toon Van Daele"
date: 
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

# Introduction

In this document we compare a selection of B-cubed biodiversity indicators between actual cube data and data from structured monitoring. For more information on both datasets check to ... document.

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE, comment = "#>")

library(targets)
Sys.setenv(TAR_PROJECT = "target_workflow")

library(b3gbi)
library(ggplot2)
```

# Biodiversity indicators

## Total occurences

### Map

```{r total occurrences map, fig.show="hold", out.width="50%"}
total_occ_map_birdcube <- tar_read(total_occ_map_cube)
plot(total_occ_map_birdcube) +
  labs(title = "Birdcube Flanders")

total_occ_map_abv <- tar_read(total_occ_map_abv)
plot(total_occ_map_abv) +
  labs(title = "ABV")
```

### Time series

```{r total occurences time series, fig.show="hold", out.width="50%"}
total_occ_ts_birdcube <- tar_read(total_occ_ts_cube)
plot(total_occ_ts_birdcube)

total_occ_ts_abv <- tar_read(total_occ_ts_abv)
plot(total_occ_ts_abv)
```

## Observed richness

### Map 1 km²

```{robserved richness map 1km, fig.show="hold", out.width="50%"}
obs_richness_map_abv_1 <- tar_read(obs_richness_map_abv_1)
plot(obs_richness_map_abv_1)

obs_richness_map_cube_1 <- tar_read(obs_richness_map_cube_1)
plot(obs_richness_map_cube_1)
```

### Map 10 km²

```{r observed richness map, fig.show="hold", out.width="50%"}
obs_richness_map_abv_10 <- tar_read(obs_richness_map_abv_10)
plot(obs_richness_map_abv_10)

obs_richness_map_cube_10 <- tar_read(obs_richness_map_cube_10)
plot(obs_richness_map_cube_10)
```

### Time series

```{r observed richness time series, fig.show="hold", out.width="50%"}
obs_richness_ts_abv <- tar_read(obs_richness_ts_abv)
plot(obs_richness_ts_abv)

obs_richness_ts_cube <- tar_read(obs_richness_ts_cube)
plot(obs_richness_ts_cube)
```

## Pielou evenness

### Map

```{r pielou evenness map, fig.show="hold", out.width="50%"}
pielou_evenness_map_abv <- tar_read(pielou_evenness_map_abv)
plot(pielou_evenness_map_abv)

pielou_evenness_map_cube <- tar_read(pielou_evenness_map_cube)
plot(pielou_evenness_map_cube)
```

### Time series

```{r pielou evenness time series, fig.show="hold", out.width="50%"}
pielou_evenness_ts_abv <- tar_read(pielou_evenness_ts_abv)
plot(pielou_evenness_ts_abv)

pielou_evenness_ts_cube <- tar_read(pielou_evenness_ts_cube)
plot(pielou_evenness_ts_cube)
```

# Species-specific indicators

We chose three species to compare the indicators between the two data sets:

-   The species with the biggest increase according to the ABV: Cetti's warbler (*Cettia cetti*)
-   The species with the biggest decrease according to the ABV: Eurasian tree sparrow (*Passer montanus*)
-   A species with a very stable trend: Common nightingale (*Luscinia megarhynchos*)

## Species occurences

### Map

### Time series

## Species range

### Map

### Time series
