---
title: "B-cubed indicators for cube vs structured data"
author: "Emma Cartuyvels, Ward Langeraert, Toon Van Daele"
date: "2024-10-31"
output: html_document
---

# Introduction

In this document we compare all B-cubed biodiversity indicators between actual cube data and data from structured monitoring.

```{r setup, include=FALSE}
library(sf)
library(rgbif)
library(dplyr)
library(b3gbi)

knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data_path <- here::here("data")
```

To start we'll read in both the biodiversity and the ABV data directly from GBIF.

```{r read data, eval=FALSE}
# nolint start
birdcubeflanders_year <- occ_download_sql(
  user = Sys.getenv("USER"),
  pwd = Sys.getenv("PSWD"),
  email = Sys.getenv("MAIL"),
  "SELECT
  \"year\",
  GBIF_EEARGCode(
      1000,
      decimalLatitude,
      decimalLongitude,
      COALESCE(coordinateUncertaintyInMeters, 1000)
    ) AS eeaCellCode,
  speciesKey,
  species,
  family, 
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,  
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
  FROM
  occurrence
  WHERE
  occurrenceStatus = 'PRESENT'
  AND NOT ARRAY_CONTAINS(issue, 'ZERO_COORDINATE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_OUT_OF_RANGE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_INVALID') 
  AND NOT ARRAY_CONTAINS(issue, 'COUNTRY_COORDINATE_MISMATCH')
  AND level1gid = 'BEL.2_1'
  AND \"year\" >= 2007
  AND \"year\" <= 2022
  AND speciesKey IS NOT NULL
  AND decimalLatitude IS NOT NULL
  AND decimalLongitude IS NOT NULL
  AND class = 'Aves'
  AND collectionCode != 'ABV'
  GROUP BY
  \"year\",
  eeaCellCode,
  speciesKey,
  family,
  species
  ORDER BY
  \"year\" ASC,
  eeaCellCode ASC,
  speciesKey ASC"
)
# nolint end
```

```{r, eval=FALSE}
occ_download_wait(birdcubeflanders_year)

birdcubeflanders <- occ_download_get(birdcubeflanders_year,
                                     path = data_path) |>
  occ_download_import()

write.csv(birdcubeflanders, paste0(data_path, "/interim/birdcubeflanders.csv"))
```

```{r, eval=FALSE}
# nolint start
abv_data_down <- occ_download_sql(
  user = Sys.getenv("USER"),
  pwd = Sys.getenv("PSWD"),
  email = Sys.getenv("MAIL"),
  "SELECT
  \"year\",
  GBIF_EEARGCode(
      1000,
      decimalLatitude,
      decimalLongitude,
      COALESCE(coordinateUncertaintyInMeters, 1000)
    ) AS eeaCellCode,
  speciesKey,
  species,
  family, 
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,  
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
  FROM
  occurrence
  WHERE 
  occurrenceStatus = 'PRESENT'
  AND NOT ARRAY_CONTAINS(issue, 'ZERO_COORDINATE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_OUT_OF_RANGE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_INVALID') 
  AND NOT ARRAY_CONTAINS(issue, 'COUNTRY_COORDINATE_MISMATCH')
  AND level1gid = 'BEL.2_1' 
  AND \"year\" >= 2007
  AND \"year\" <= 2022
  AND speciesKey IS NOT NULL
  AND decimalLatitude IS NOT NULL
  AND decimalLongitude IS NOT NULL
  AND class = 'Aves'
  AND collectionCode = 'ABV'
  GROUP BY
  \"year\",
  eeaCellCode,
  speciesKey,
  family,
  species
  ORDER BY
  \"year\" ASC,
  eeaCellCode ASC,
  speciesKey ASC"
)
# nolint end
```

```{r, eval=FALSE}
occ_download_wait(abv_data_down)

abv_data <- occ_download_get(abv_data_down,
                             path = data_path) |>
  occ_download_import()

write.csv(abv_data, "../data/interim/abv_data.csv")
```

```{r}
birdcubeflanders <- read.csv("../data/interim/birdcubeflanders.csv")

birdcube <- process_cube(birdcubeflanders,
                         cols_occurrences = "n")
```

```{r}
abv_data <- read.csv("../data/interim/abv_data.csv")

abv <- process_cube(abv_data,
                    cols_occurrences = "n")
```

# General Biodiversity indicators

## Observed richness

### Map
```{r observed richness 1km map, fig.show="hold", out.width="50%"}
obs_richness_map_birdcube <- obs_richness_map(birdcube, cell_size = 1)
plot(obs_richness_map_birdcube)

obs_richness_map_abv <- obs_richness_map(abv, cell_size = 1)
plot(obs_richness_map_abv)
```

```{r observed richness 10km map, fig.show="hold", out.width="50%"}
obs_richness_map_birdcube <- obs_richness_map(birdcube, cell_size = 10)
plot(obs_richness_map_birdcube)

obs_richness_map_abv <- obs_richness_map(abv, cell_size = 10)
plot(obs_richness_map_abv)
```
These maps, as expected, show clear differences in the species richness observed, we also see no similar pattern in where higher species richness is seen.

### Time series

```{r observed richness over time, fig.show="hold", out.width="50%"}
obs_richness_ts_birdcube <- obs_richness_ts(birdcube)
plot(obs_richness_ts_birdcube)

obs_richness_ts_abv <- obs_richness_ts(abv)
plot(obs_richness_ts_abv)
```
## Total occurrences

### Map

```{r total occurrences map, fig.show="hold", out.width="50%"}
total_occ_map_birdcube <- total_occ_map(birdcube, cell_size = 10)
plot(total_occ_map_birdcube)

total_occ_map_abv <- total_occ_map(abv, cell_size = 10)
plot(total_occ_map_abv)
```

### Time series

```{r total occurences time series, fig.show="hold", out.width="50%"}
total_occ_ts_birdcube <- total_occ_ts(birdcube)
plot(total_occ_ts_birdcube)

total_occ_ts_abv <- total_occ_ts(abv)
plot(total_occ_ts_abv)
```

## Pielou evenness

### Map

```{r pielou evenness map, fig.show="hold", out.width="50%"}
pielou_evenness_map_birdcube <- pielou_evenness_map(birdcube, cell_size = 10)
plot(pielou_evenness_map_birdcube)

pielou_evenness_map_abv <- pielou_evenness_map(abv, cell_size = 10)
plot(pielou_evenness_map_abv)
```

### Time series

```{r pielou evenness time series, fig.show="hold", out.width="50%"}
pielou_evenness_ts_birdcube <- pielou_evenness_ts(birdcube)
plot(pielou_evenness_ts_birdcube)

pielou_evenness_ts_abv <- pielou_evenness_ts(abv)
plot(pielou_evenness_ts_abv)
```

## Williams' Evenness

### Map

```{r williams evenness map, fig.show="hold", out.width="50%"}
williams_evenness_map_birdcube <- williams_evenness_map(birdcube,
                                                        cell_size = 10)
plot(williams_evenness_map_birdcube)

williams_evenness_map_abv <- williams_evenness_map(abv, cell_size = 10)
plot(williams_evenness_map_abv)
```

### Time series

```{r williams evenness time series, fig.show="hold", out.width="50%"}
williams_evenness_ts_birdcube <- williams_evenness_ts(birdcube)
plot(williams_evenness_ts_birdcube)

williams_evenness_ts_abv <- williams_evenness_ts(abv)
plot(williams_evenness_ts_abv)
```

## Cumulative Species Richness

### Time series

```{r cumulative species richness over time, fig.show="hold", out.width="50%"}
cum_richness_ts_birdcube <- cum_richness_ts(birdcube)
plot(cum_richness_ts_birdcube)

cum_richness_ts_abv <- cum_richness_ts(abv)
plot(cum_richness_ts_abv)
```

## Density of Occurrences

### Map

```{r density of occurences map, fig.show="hold", out.width="50%"}
occ_density_map_birdcube <- occ_density_map(birdcube, cell_size = 10)
plot(occ_density_map_birdcube)

occ_density_map_abv <- occ_density_map(abv, cell_size = 10)
plot(occ_density_map_abv)
```

### Time series

```{r density of occurences time series, fig.show="hold", out.width="50%"}
occ_density_ts_birdcube <- occ_density_ts(birdcube)
plot(occ_density_ts_birdcube)

occ_density_ts_abv <- occ_density_ts(abv)
plot(occ_density_ts_abv)
```

## Abundance-Based Rarity

### Map

```{r abundance-based rarity map, fig.show="hold", out.width="50%"}
ab_rarity_map_birdcube <- ab_rarity_map(birdcube, cell_size = 10)
plot(ab_rarity_map_birdcube)

ab_rarity_map_abv <- ab_rarity_map(abv, cell_size = 10)
plot(ab_rarity_map_abv)
```

### Time series

```{r abundance-based rarity time series, fig.show="hold", out.width="50%"}
ab_rarity_ts_birdcube <- ab_rarity_ts(birdcube)
plot(ab_rarity_ts_birdcube)

ab_rarity_ts_abv <- ab_rarity_ts(abv)
plot(ab_rarity_ts_abv)
```

## Area-Based Rarity

### Map

```{r area-based rarity map, fig.show="hold", out.width="50%"}
area_rarity_map_birdcube <- area_rarity_map(birdcube, cell_size = 10)
plot(area_rarity_map_birdcube)

area_rarity_map_abv <- area_rarity_map(abv, cell_size = 10)
plot(area_rarity_map_abv)
```

### Time series

```{r area-based rarity time series, fig.show="hold", out.width="50%"}
area_rarity_ts_birdcube <- area_rarity_ts(birdcube)
plot(area_rarity_ts_birdcube)

area_rarity_ts_abv <- area_rarity_ts(abv)
plot(area_rarity_ts_abv)
```

## Mean Year of Occurrence

### Map

```{r newness map, fig.show="hold", out.width="50%"}
newness_map_birdcube <- newness_map(birdcube, cell_size = 10)
plot(newness_map_birdcube)

newness_map_abv <- newness_map(abv, cell_size = 10)
plot(newness_map_abv)
```

### Time series

```{r newness time series, fig.show="hold", out.width="50%"}
newness_ts_birdcube <- newness_ts(birdcube)
plot(newness_ts_birdcube)

newness_ts_abv <- newness_ts(abv)
plot(newness_ts_abv)
```

## Taxonomic Distinctness

### Map

```{r taxonomic distinctness map, fig.show="hold", out.width="50%"}
tax_distinct_map_birdcube <- tax_distinct_map(birdcube, cell_size = 10,
                                              check = TRUE)
plot(tax_distinct_map_birdcube)

tax_distinct_map_abv <- tax_distinct_map(abv, cell_size = 10, check = TRUE)
plot(tax_distinct_map_abv)
```

### Time series

```{r taxonomic distinctness time series, eval = FALSE, fig.show="hold", out.width="50%"}
tax_distinct_ts_birdcube <- tax_distinct_ts(birdcube)
plot(tax_distinct_ts_birdcube)

tax_distinct_ts_abv <- tax_distinct_ts(abv, check = TRUE)
plot(tax_distinct_ts_abv)
```

## Species Richness (Estimated by Coverage-Based Rarefaction)

### Map

```{r species richness map, eval = FALSE, fig.show="hold", out.width="50%"}
hill0_map_birdcube <- hill0_map(birdcube, cell_size = 10)
plot(hill0_map_birdcube)

hill0_map_abv <- hill0_map(abv, cell_size = 10)
plot(hill0_map_abv)
```

### Time series

```{r species richness time series, fig.show="hold", out.width="50%"}
hill0_ts_birdcube <- hill0_ts(birdcube)
plot(hill0_ts_birdcube)

hill0_ts_abv <- hill0_ts(abv)
plot(hill0_ts_abv)
```

## Hill-Shannon Diversity (Estimated by Coverage-Based Rarefaction)

### Map

```{r hill-shannnon diversity map, eval = FALSE, fig.show="hold", out.width="50%"}
hill1_map_birdcube <- hill1_map(birdcube, cell_size = 10)
plot(hill1_map_birdcube)

hill1_map_abv <- hill1_map(abv, cell_size = 10)
plot(hill1_map_abv)
```

### Time series

```{r hill-shannon diversity time series, fig.show="hold", out.width="50%"}
hill1_ts_birdcube <- hill1_ts(birdcube)
plot(hill1_ts_birdcube)

hill1_ts_abv <- hill1_ts(abv)
plot(hill1_ts_abv)
```

## Hill-Simpson Diversity (Estimated by Coverage-Based Rarefaction)

### Map

```{r hill-simpson diversity map, eval=FALSE, fig.show="hold", out.width="50%"}
hill2_map_birdcube <- hill2_map(birdcube, cell_size = 10)
plot(hill2_map_birdcube)

hill2_map_abv <- hill2_map(abv, cell_size = 10)
plot(hill2_map_abv)
```

### Time series

```{r hill-simpson diversity time series, fig.show="hold", out.width="50%"}
hill2_ts_birdcube <- hill2_ts(birdcube)
plot(hill2_ts_birdcube)

hill2_ts_abv <- hill2_ts(abv)
plot(hill2_ts_abv)
```

## Occupancy Turnover

### Time series

```{r occupancy turnover time series, fig.show="hold", out.width="50%"}
occ_turnover_ts_birdcube <- occ_turnover_ts(birdcube)
plot(occ_turnover_ts_birdcube)

occ_turnover_ts_abv <- occ_turnover_ts(abv)
plot(occ_turnover_ts_abv)
```


# Species-specific Biodiversity indicators

To compare these indicators we selected three species: the extremely common ..., generalist ..., the rare species ... (work with tabs possibly)

## Species occurrences

## Species range

Ferro and Flick (2015) find that you need 15 collections of a specimen to predict it's distribution.

### Map

```{r}
spec_range_map_birdcube <- spec_range_map(birdcube)
plot(spec_range_map_birdcube, species = "Luscinia svecica")
```

### Time series

```{r}
spec_range_ts_birdcube <- spec_range_ts(birdcube)
plot(spec_range_ts_birdcube, species = "Luscinia svecica")
```




