---
title: "Prepare ABV data"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
# Load packages
library(tidyverse) # Data wrangling and visualisation
library(zen4R)     # Download from zenodo
library(here)      # Relative paths
library(sf)        # Work with spatial data
```

# Goal

Load and save structured data of the “Common Breeding Bird Survey Flanders” (ABV).
Load and save unstructured data.

# Structured data
## Sampling framework

We download the sampling framework from zenodo.

```{r}
# Data path and create directory if necessary
data_path <- here::here("data", "raw")
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)

# Download data from zenodo if necessary
file_sampling_framework <- file.path(
  data_path,
  "steekproefkader.csv")

if (file.exists(file_sampling_framework)) {
  sampling_framework_abv <- read_csv(file_sampling_framework,
                                     show_col_types = FALSE)
} else {
  download_zenodo(
    doi = "10.5281/zenodo.10103472",
    path = data_path,
    files = list("steekproefkader.csv"))

  sampling_framework_abv <- read_csv(file_sampling_framework,
                                     show_col_types = FALSE)
}
```

We get a dataset with land use category per UTM grid cell (1x1 km).

```{r}
# Explore dataframe
head(sampling_framework_abv)
```

## Occurrence data

Occurrence data is downloaded from GBIF.org.
Unzip and store folders under *./data/raw*.

**ABV - Common breeding birds in Flanders, Belgium**: 0000102-240626123714530.zip

GBIF.org (26 June 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.w8va93

**ABV - Common breeding birds in Flanders, Belgium (post 2016)**: 0000105-240626123714530.zip

GBIF.org (26 June 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.4htaqh

We read in these datasets.

```{r}
# Read ABV data as tab delimited
abv_data_raw <- read_delim(file.path(data_path,
                                     "0000102-240626123714530",
                                     "occurrence.txt"),
                   delim = "\t",
                   show_col_types = FALSE)
abv_data2_raw <- read_delim(file.path(data_path,
                                      "0000105-240626123714530",
                                      "occurrence.txt"),
                   delim = "\t",
                   show_col_types = FALSE)
```

We combine the datasets.

```{r}
# Combine to single dataframe
abv_data_total_raw <- bind_rows(abv_data_raw, abv_data2_raw)

# Remove empty columns
abv_data_total <- abv_data_total_raw[, colSums(is.na(abv_data_total_raw)) <
                                       nrow(abv_data_total_raw)]
```

We get a big dataframe with all occurrences.

```{r}
# Explore dataframe
glimpse(abv_data_total)
```

# Unstructured data

We retrieved species occurrence cubes using Microsoft Azure Databricks as described [here](https://techdocs.gbif.org/en/data-use/b-cubed/generate-cube-databricks).

We first selected all bird data in Belgium `birds_data.belgium` from 2007-2022 using this query:

```sql
CREATE DATABASE IF NOT EXISTS birds_data;
CREATE TABLE IF NOT EXISTS birds_data.belgium AS SELECT
  *
FROM
  gbif.occurrence
WHERE
  occurrenceStatus = 'PRESENT'
  AND NOT array_contains(issue.array_element, 'ZERO_COORDINATE')
  AND NOT array_contains(issue.array_element, 'COORDINATE_OUT_OF_RANGE')
  AND NOT array_contains(issue.array_element, 'COORDINATE_INVALID')
  AND NOT array_contains(issue.array_element, 'COUNTRY_COORDINATE_MISMATCH')
  AND countrycode = 'BE'
  AND year BETWEEN 2007 AND 2022
  AND speciesKey IS NOT NULL 
  AND decimallatitude IS NOT NULL
  AND decimallongitude IS NOT NULL
  AND class = 'Aves';
```

We excluded ABV data into a new table `birds_data.unstructured` using this SQL code:

```sql
CREATE TABLE IF NOT EXISTS birds_data.unstructured AS SELECT
  *
FROM
  birds_data.belgium
WHERE
  collectionCode != 'ABV';
```

## Occurrence cube: MGRS per month

We generate the occurrence cube on a 1 km scale per month.
To account for sampling bias, the specification recommends including an occurrence count for a higher taxon, typically the family.

```sql
SELECT
  -- Dimensions
  CONCAT_WS('-', year, month) AS yearMonth,
  mgrsCode(
    1000,
    decimalLatitude,
    decimalLongitude,
    COALESCE(coordinateUncertaintyInMeters, 1000)
  ) AS mgrsCode,
  speciesKey,
  species,
  family,
  -- Measurements
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,
  -- Higher taxon measurement
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
FROM
  birds_data.unstructured
WHERE month IS NOT NULL
GROUP BY
  yearMonth,
  mgrsCode,
  speciesKey,
  family,
  species
ORDER BY
  yearMonth ASC,
  mgrsCode ASC,
  speciesKey ASC;
```

The contributions of GBIF-mediated datasets can be determined with a second SQL query, using the WHERE clause of the first. In effect, this is a two-dimensional "cube".

```sql
SELECT
  -- Dimensions
  datasetKey,
  collectionCode,
  license,
  -- Measurement
  COUNT(*) AS n
FROM
  birds_data.unstructured
WHERE month IS NOT NULL
GROUP BY
  datasetKey,
  collectionCode,
  license;
```

We read in the datacube.

```{r}
# Read in datacube from CSV
birdcubeflanders_yearmonth <- read_csv(
  file.path(data_path, "birdcubeflanders_yearmonth_mgrs.csv"),
  show_col_types = FALSE)

# Explore dataframe
glimpse(birdcubeflanders_yearmonth)
```

## Occurrence cube: MGRS per year

We generate the occurrence cube on a 1 km scale per year similar as above.

```sql
SELECT
  -- Dimensions
  year,
  mgrsCode(
    1000,
    decimalLatitude,
    decimalLongitude,
    COALESCE(coordinateUncertaintyInMeters, 1000)
  ) AS mgrsCode,
  speciesKey,
  species,
  family,
  -- Measurements
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,
  -- Higher taxon measurement
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
FROM
  birds_data.unstructured
GROUP BY
  year,
  mgrsCode,
  speciesKey,
  family,
  species
ORDER BY
  year ASC,
  mgrsCode ASC,
  speciesKey ASC;
```

The contributions of GBIF-mediated datasets:

```sql
SELECT
  -- Dimensions
  datasetKey,
  collectionCode,
  license,
  -- Measurement
  COUNT(*) AS n
FROM
  birds_data.unstructured
GROUP BY
  datasetKey,
  collectionCode,
  license;
```

We read in the datacube.

```{r}
# Read in datacube from CSV
birdcubeflanders_year <- read_csv(
  file.path(data_path, "birdcubeflanders_year_mgrs.csv"),
  show_col_types = FALSE)

# Explore dataframe
glimpse(birdcubeflanders_year)
```

# Select Flanders grid cells

The datacubes cover multiple zones although Flanders is present only in zone 31U. 

```{r}
# Number of rows per zone
table(substring(birdcubeflanders_year$mgrsCode, 1, 3))
```

We add 31U to the tag names of the sampling framework and the ABV data.

```{r}
# Create new column with correct MGRS code
sampling_framework_abv <- sampling_framework_abv %>%
  mutate(mgrsCode = paste0("31U", UTM))

abv_data_total <- abv_data_total %>%
  mutate(verbatimLocality2 = gsub("ABV_", "", verbatimLocality)) %>%
  mutate(mgrsCode = paste0("31U", verbatimLocality2))
```

We load in the UTM grid for Flanders (1 km) and add 31U to the tag names.

```{r}
# Read UTM 1 km grid and add new column with correct MGRS code
utm_grid <- read_sf(file.path(data_path, "utm_grid", "utm1_vl.shp"))
utm_grid <- utm_grid %>%
  mutate(mgrsCode = paste0("31U", TAG))

# Explore dataframe
glimpse(utm_grid)
```

We add the geometry to the data layers by taking an inner join.

```{r}
# Add UTM geometry by taking an inner join
sampling_framework_abv_sf <- utm_grid %>%
  inner_join(sampling_framework_abv, by = join_by(mgrsCode)) %>%
  st_sf(sf_column_name = "geometry")

# Visualise spatial distribution sampling framework
utm_grid %>%
  left_join(sampling_framework_abv, by = join_by(mgrsCode)) %>%
  st_sf(sf_column_name = "geometry") %>%
  ggplot() +
    geom_sf(aes(fill = Stratum),
            col = alpha("white", 0)) +
    ggtitle("UTM 1 km grid cells Flanders")
```

```{r}
# Add UTM geometry by taking an inner join
abv_data_total_sf <- utm_grid %>%
  inner_join(abv_data_total, by = join_by(mgrsCode)) %>%
  st_sf(sf_column_name = "geometry")

# Visualise spatial distribution of the ABV data
utm_grid %>%
  left_join(abv_data_total %>%
              filter(occurrenceStatus == "PRESENT") %>%
              distinct(occurrenceStatus, mgrsCode),
            by = join_by(mgrsCode)) %>%
  ggplot() +
    geom_sf(aes(fill = occurrenceStatus), col = alpha("white", 0)) +
    ggtitle("UTM 1 km grid cells Flanders")
```

We select cube data from Flanders and add the geometry to the data layers by taking an inner join.
Per month:

```{r}
# Add UTM geometry and select data by taking an inner join
birdcubeflanders_yearmonth_sf <- utm_grid %>%
  inner_join(birdcubeflanders_yearmonth, by = join_by(mgrsCode)) %>%
  st_sf(sf_column_name = "geometry")
```

```{r}
# Visualise spatial distribution data cube as number of species
utm_grid %>%
  left_join(birdcubeflanders_yearmonth %>%
              group_by(mgrsCode) %>%
              summarise(n_species = n_distinct(species), .groups = "drop"),
            by = join_by(mgrsCode)) %>%
  ggplot() +
    geom_sf(aes(fill = n_species), col = alpha("white", 0)) +
    scale_fill_viridis_c(option = "inferno") +
    ggtitle("Bird cube data from Flanders per month")
```

Per year:

```{r}
# Add UTM geometry and select data by taking an inner join
birdcubeflanders_year_sf <- utm_grid %>%
  inner_join(birdcubeflanders_year, by = join_by(mgrsCode)) %>%
  st_sf(sf_column_name = "geometry")
```

```{r}
# Visualise spatial distribution data cube as number of species
utm_grid %>%
  left_join(birdcubeflanders_year %>%
              group_by(mgrsCode) %>%
              summarise(n_species = n_distinct(species), .groups = "drop"),
            by = join_by(mgrsCode)) %>%
  ggplot() +
    geom_sf(aes(fill = n_species), col = alpha("white", 0)) +
    scale_fill_viridis_c(option = "inferno") +
    ggtitle("Bird cube data from Flanders per year")
```

We see a striped pattern because of large uncertainty.
We filter `minCoordinateUncertaintyInMeters` smaller or equal to 1000 meters:

```{r}
# Add filter
utm_grid %>%
  left_join(birdcubeflanders_yearmonth %>%
              filter(minCoordinateUncertaintyInMeters <= 1000) %>%
              group_by(mgrsCode) %>%
              summarise(n_species = n_distinct(species), .groups = "drop"),
            by = join_by(mgrsCode)) %>%
  ggplot() +
    geom_sf(aes(fill = n_species), col = alpha("white", 0)) +
    scale_fill_viridis_c(option = "inferno") +
    ggtitle("Bird cube data from Flanders per month")
```

# Write out data

We write out the data for exploration and analysis.

```{r}
out_path <- here::here("data", "interim")
dir.create(out_path, showWarnings = FALSE, recursive = TRUE)

# Structured data
write_sf(abv_data_total_sf,
         file.path(out_path, "abv_data_total.gpkg"))
write_sf(sampling_framework_abv_sf,
         file.path(out_path, "sampling_framework_abv.gpkg"))

# Unstructured data
write_sf(birdcubeflanders_yearmonth_sf,
         file.path(out_path, "birdcubeflanders_yearmonth.gpkg"))
write_sf(birdcubeflanders_year_sf,
         file.path(out_path, "birdcubeflanders_year.gpkg"))
```
