---
title: "Download and prepare ABV data"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
# Load packages
library(tidyverse) # Data wrangling and visualisation
library(zen4R)     # Download from zenodo
library(here)      # Relative paths
library(sf)        # Work with spatial data

# Source
source(here("source/R/download_occ_cube.R"))
```

# Goal

Load and save structured data of the “Common Breeding Bird Survey Flanders” (ABV).
Load and save unstructured data.

# Structured data
## Sampling framework

We download the sampling framework from zenodo.

```{r}
# Data path and create directory if necessary
data_path <- here("data", "raw")
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)

# Download data from zenodo if necessary
file_sampling_framework <- file.path(
  data_path,
  "steekproefkader.csv"
)

if (file.exists(file_sampling_framework)) {
  sampling_framework_abv <- read_csv(file_sampling_framework,
                                     show_col_types = FALSE)
} else {
  download_zenodo(
    doi = "10.5281/zenodo.10103472",
    path = data_path,
    files = list("steekproefkader.csv")
  )

  sampling_framework_abv <- read_csv(file_sampling_framework,
                                     show_col_types = FALSE)
}
```

We get a dataset with land use category per UTM grid cell (1x1 km).

```{r}
# Explore dataframe
head(sampling_framework_abv)
```

## Occurrence data

The ABV data is downloaded as a cube from GBIF.org. We set the randomization to zero, since the monitoring takes place in specific squares.
The zip file is stored under *./data/raw*.

> GBIF.org (15 April 2025) GBIF Occurrence Download  https://doi.org/10.15468/dl.ddzhrc

```{r}
# nolint start: line_length_linter.
query_abv <- "SELECT
  \"year\",
  GBIF_MGRSCode(1000, decimalLatitude, decimalLongitude, 0) AS mgrsCode,
  speciesKey,
  species,
  family,
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 0)) AS minCoordinateUncertaintyInMeters,
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
  FROM
  occurrence
  WHERE
  occurrenceStatus = 'PRESENT'
  AND NOT occurrence.basisofrecord IN ('FOSSIL_SPECIMEN', 'LIVING_SPECIMEN')
  AND NOT ARRAY_CONTAINS(issue, 'ZERO_COORDINATE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_OUT_OF_RANGE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_INVALID')
  AND NOT ARRAY_CONTAINS(issue, 'COUNTRY_COORDINATE_MISMATCH')
  AND level1gid = 'BEL.2_1'
  AND \"year\" >= 2007
  AND \"year\" <= 2022
  AND speciesKey IS NOT NULL
  AND decimalLatitude IS NOT NULL
  AND decimalLongitude IS NOT NULL
  AND class = 'Aves'
  AND collectionCode = 'ABV'
  GROUP BY
  \"year\",
  mgrsCode,
  speciesKey,
  family,
  species
  ORDER BY
  \"year\" ASC,
  mgrsCode ASC,
  speciesKey ASC"
# nolint end

abv_data_total <- download_occ_cube(
  sql_query = query_abv,
  file = "abv_data.csv",
  path = data_path,
  overwrite = FALSE
)
```

We get a big dataframe with all occurrences.

```{r}
# Explore dataframe
glimpse(abv_data_total)
```

# Unstructured data

The cube data is downloaded as a cube from GBIF.org.
The zip file is stored under *./data/raw*.

> GBIF.org (15 April 2025) GBIF Occurrence Download  https://doi.org/10.15468/dl.mn4ybb

```{r}
# nolint start: line_length_linter.
query_birdcube <- "SELECT
  \"year\",
  GBIF_MGRSCode(1000, decimalLatitude, decimalLongitude,
  COALESCE(coordinateUncertaintyInMeters, 1000)) AS mgrsCode,
  speciesKey,
  species,
  family,
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
  FROM
  occurrence
  WHERE
  occurrenceStatus = 'PRESENT'
  AND NOT occurrence.basisofrecord IN ('FOSSIL_SPECIMEN', 'LIVING_SPECIMEN')
  AND NOT ARRAY_CONTAINS(issue, 'ZERO_COORDINATE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_OUT_OF_RANGE')
  AND NOT ARRAY_CONTAINS(issue, 'COORDINATE_INVALID')
  AND NOT ARRAY_CONTAINS(issue, 'COUNTRY_COORDINATE_MISMATCH')
  AND level1gid = 'BEL.2_1'
  AND \"year\" >= 2007
  AND \"year\" <= 2022
  AND speciesKey IS NOT NULL
  AND decimalLatitude IS NOT NULL
  AND decimalLongitude IS NOT NULL
  AND class = 'Aves'
  AND collectionCode != 'ABV'
  GROUP BY
  \"year\",
  mgrsCode,
  speciesKey,
  family,
  species
  ORDER BY
  \"year\" ASC,
  mgrsCode ASC,
  speciesKey ASC"
# nolint end

birdcube_data_total <- download_occ_cube(
  sql_query = query_birdcube,
  file = "birdcube_data.csv",
  path = data_path,
  overwrite = FALSE
)
```

We get a big dataframe with all occurrences.

```{r}
# Explore dataframe
glimpse(birdcube_data_total)
```

# Select Flanders grid cells
The datacubes cover multiple zones although Flanders is present only in zone 31U. 

```{r}
# Number of rows per zone
table(substring(abv_data_total$mgrscode, 1, 3))
```

We add 31U to the tag names of the sampling framework.

```{r}
# Create new column with correct MGRS code
sampling_framework_abv <- sampling_framework_abv %>%
  mutate(mgrscode = paste0("31U", UTM))
```

We load in the UTM grid for Flanders (1 km) and add 31U to the tag names.

```{r}
# Read UTM 1 km grid and add new column with correct MGRS code
utm_grid <- read_sf(file.path(data_path, "utm_grid", "utm1_vl.shp"))
utm_grid <- utm_grid %>%
  mutate(mgrscode = paste0("31U", TAG))

# Explore dataframe
glimpse(utm_grid)
```

We add the geometry to the data layers by taking an inner join.

```{r}
# Add UTM geometry by taking an inner join
sampling_framework_abv_sf <- utm_grid %>%
  inner_join(sampling_framework_abv, by = join_by(mgrscode)) %>%
  st_sf(sf_column_name = "geometry")

# Visualise spatial distribution sampling framework
utm_grid %>%
  left_join(sampling_framework_abv, by = join_by(mgrscode)) %>%
  st_sf(sf_column_name = "geometry") %>%
  ggplot() +
  geom_sf(aes(fill = Stratum),
          col = alpha("white", 0)) +
  ggtitle("Sampling framework ABV")
```

```{r}
# Add UTM geometry by taking an inner join
abv_data_total_sf <- utm_grid %>%
  inner_join(abv_data_total, by = join_by(mgrscode)) %>%
  st_sf(sf_column_name = "geometry")

# Visualise spatial distribution of the ABV data
utm_grid %>%
  left_join(abv_data_total %>%
              mutate(presence = "present") %>%
              distinct(presence, mgrscode),
            by = join_by(mgrscode)) %>%
  mutate(sample = ifelse(is.na(presence), "no", "yes")) %>%
  ggplot() +
  geom_sf(aes(fill = sample), col = alpha("white", 0)) +
  ggtitle("ABV sampling locations")
```

We select cube data from Flanders and add the geometry to the data layers by taking an inner join.

```{r}
# Add UTM geometry and select data by taking an inner join
birdcube_data_total_sf <- utm_grid %>%
  inner_join(birdcube_data_total, by = join_by(mgrscode)) %>%
  st_sf(sf_column_name = "geometry")
```

```{r}
# Visualise spatial distribution data cube as number of species
utm_grid %>%
  left_join(birdcube_data_total %>%
              group_by(mgrscode) %>%
              summarise(n_species = n_distinct(species), .groups = "drop"),
            by = join_by(mgrscode)) %>%
  ggplot() +
  geom_sf(aes(fill = n_species), col = alpha("white", 0)) +
  scale_fill_viridis_c(option = "inferno") +
  ggtitle("Bird cube data from Flanders")
```

We see a dotted pattern because of large uncertainty.
We filter `minCoordinateUncertaintyInMeters` smaller or equal to 1000 meters:

```{r}
# Add filter
utm_grid %>%
  left_join(birdcube_data_total %>%
              filter(mincoordinateuncertaintyinmeters <= 1000) %>%
              group_by(mgrscode) %>%
              summarise(n_species = n_distinct(species), .groups = "drop"),
            by = join_by(mgrscode)) %>%
  ggplot() +
  geom_sf(aes(fill = n_species), col = alpha("white", 0)) +
  scale_fill_viridis_c(option = "inferno") +
  ggtitle("Bird cube data from Flanders")
```

# Correction of species names

There are some double accepted species names that cause trouble.

```{r}
abv_data_total_sf <- abv_data_total_sf %>%
  mutate(
    species = case_when(
      species == "Dendrocopus major" ~ "Dendrocopos major",
      species == "Saxicola torquatus" ~ "Saxicola rubicola",
      TRUE ~ species
    ),
    specieskey = case_when(
      species == "Dendrocopos major" ~ 2477968,
      species == "Saxicola rubicola" ~ 4408759,
      TRUE ~ specieskey
    )
  )
```

```{r}
birdcube_data_total_sf <- birdcube_data_total_sf %>%
  mutate(
    species = case_when(
      species == "Poecile montanus" ~ "Parus montanus",
      TRUE ~ species
    ),
    specieskey = case_when(
      species == "Parus montanus" ~ 4409010,
      TRUE ~ specieskey
    )
  )
```

# Write out data

We select the columns we want in a logical order:

```{r}
abv_data_out_sf <- abv_data_total_sf %>%
  select("mgrscode", "year", "specieskey", "species", "family", "n",
         "mincoordinateuncertaintyinmeters", "familycount", "geometry")
abv_data_out <- st_drop_geometry(abv_data_out_sf)

colnames(sampling_framework_abv_sf) <- tolower(
  colnames(sampling_framework_abv_sf)
)
sampling_framework_out_sf <- sampling_framework_abv_sf %>%
  select("mgrscode", "stratum", "oppervlakte", "landbouw", "urbaan", "bos",
         "suburbaan", "heideduin", "moeraswater", "geometry")
sampling_framework_out <- st_drop_geometry(sampling_framework_out_sf)

birdcube_data_out_sf <- birdcube_data_total_sf %>%
  select("mgrscode", "year", "specieskey", "species", "family", "n",
         "mincoordinateuncertaintyinmeters", "familycount", "geometry")
birdcube_data_out <- st_drop_geometry(birdcube_data_out_sf)
```

We write out the data for exploration and analysis.

```{r}
out_path <- here("data", "interim")
dir.create(out_path, showWarnings = FALSE, recursive = TRUE)

# Structured data
## CSV
write_csv(abv_data_out,
          file.path(out_path, "abv_data_cube_1km.csv"))
write_csv(sampling_framework_out,
          file.path(out_path, "sampling_framework_abv.csv"))

## Spatial object
write_sf(abv_data_out_sf,
         file.path(out_path, "abv_data_cube_1km.gpkg"))
write_sf(sampling_framework_out_sf,
         file.path(out_path, "sampling_framework_abv.gpkg"))

# Unstructured data
## CSV
write_csv(birdcube_data_out,
          file.path(out_path, "birdflanders_cube_1km.csv"))

## Spatial object
write_sf(birdcube_data_out_sf,
         file.path(out_path, "birdflanders_cube_1km.gpkg"))
```
