---
title: "Prepare ABV data"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(zen4R)
library(here)
```

# Goal

Load and save structured data of the “Common Breeding Bird Survey Flanders” (ABV).
Load and save unstructured data.

# Structured data
## Sampling framework

We download the sampling framework from zenodo.

```{r}
# Data path and create directory if necessary
data_path <- here::here("data", "raw")
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)

# Download data from zenodo if necessary
file_sampling_framework <- file.path(
  data_path,
  "steekproefkader.csv")

if (file.exists(file_sampling_framework)) {
  sampling_framework_abv <- read_csv(file_sampling_framework,
                                     show_col_types = FALSE)
} else {
  download_zenodo(
    doi = "10.5281/zenodo.10103472",
    path = data_path,
    files = list("steekproefkader.csv"))
  
  sampling_framework_abv <- read_csv(file_sampling_framework,
                                     show_col_types = FALSE)
}
```

We get a dataset with land use category per UTM grid cell (1x1 km).

```{r}
head(sampling_framework_abv)
```

## Occurrence data

Occurrence data is downloaded from GBIF.org.
Unzip and store folders under *./data/raw*.

**ABV - Common breeding birds in Flanders, Belgium**: 0000102-240626123714530.zip

GBIF.org (26 June 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.w8va93

**ABV - Common breeding birds in Flanders, Belgium (post 2016)**: 0000105-240626123714530.zip

GBIF.org (26 June 2024) GBIF Occurrence Download  https://doi.org/10.15468/dl.4htaqh

We read in these datasets.

```{r}
abv_data_raw <- read_delim(file.path(data_path,
                                     "0000102-240626123714530",
                                     "occurrence.txt"),
                   delim = "\t",
                   show_col_types = FALSE)
abv_data2_raw <- read_delim(file.path(data_path,
                                      "0000105-240626123714530",
                                      "occurrence.txt"),
                   delim = "\t",
                   show_col_types = FALSE)
```

We combine the datasets.

```{r}
abv_data_total_raw <- bind_rows(abv_data_raw, abv_data2_raw)
  
# Remove empty columns
abv_data_total <- abv_data_total_raw[, colSums(is.na(abv_data_total_raw)) < 
                                       nrow(abv_data_total_raw)]
```

We get a big dataframe with all occurrences.

```{r}
glimpse(abv_data_total)
```

# Unstructured data

We retrieved species occurrence cubes using Microsoft Azure Databricks as described [here](https://techdocs.gbif.org/en/data-use/b-cubed/generate-cube-databricks).

We first selected all bird data in Belgium `birds_data.belgium` from 2007-2022 using this query:

```sql
CREATE DATABASE IF NOT EXISTS birds_data;
CREATE TABLE IF NOT EXISTS birds_data.belgium AS SELECT
  *
FROM
  gbif.occurrence
WHERE
  occurrenceStatus = 'PRESENT'
  AND NOT array_contains(issue.array_element, 'ZERO_COORDINATE')
  AND NOT array_contains(issue.array_element, 'COORDINATE_OUT_OF_RANGE')
  AND NOT array_contains(issue.array_element, 'COORDINATE_INVALID')
  AND NOT array_contains(issue.array_element, 'COUNTRY_COORDINATE_MISMATCH')
  AND countrycode = 'BE'
  AND year BETWEEN 2007 AND 2022
  AND speciesKey IS NOT NULL 
  AND decimallatitude IS NOT NULL
  AND decimallongitude IS NOT NULL
  AND class = 'Aves';
```

We excluded ABV data into a new table `birds_data.unstructured` using this SQL code:

```sql
CREATE TABLE IF NOT EXISTS birds_data.unstructured AS SELECT
  *
FROM
  birds_data.belgium
WHERE
  collectionCode != 'ABV';
```

## Occurrence cube: MGRS per month

We generate the occurrence cube on a 1 km scale per month.
To account for sampling bias, the specification recommends including an occurrence count for a higher taxon, typically the family.

```sql
SELECT
  -- Dimensions
  CONCAT_WS('-', year, month) AS yearMonth,
  mgrsCode(
    1000,
    decimalLatitude,
    decimalLongitude,
    COALESCE(coordinateUncertaintyInMeters, 1000)
  ) AS mgrsCode,
  speciesKey,
  species,
  family,
  -- Measurements
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,
  -- Higher taxon measurement
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
FROM
  birds_data.unstructured
WHERE month IS NOT NULL
GROUP BY
  yearMonth,
  mgrsCode,
  speciesKey,
  family,
  species
ORDER BY
  yearMonth ASC,
  mgrsCode ASC,
  speciesKey ASC;
```

The contributions of GBIF-mediated datasets can be determined with a second SQL query, using the WHERE clause of the first. In effect, this is a two-dimensional "cube".

```sql
SELECT
  -- Dimensions
  datasetKey,
  collectionCode,
  license,
  -- Measurement
  COUNT(*) AS n
FROM
  birds_data.unstructured
WHERE month IS NOT NULL
GROUP BY
  datasetKey,
  collectionCode,
  license;
```

## Occurrence cube: MGRS per year

We generate the occurrence cube on a 1 km scale per year similar as above.

```sql
SELECT
  -- Dimensions
  year,
  mgrsCode(
    1000,
    decimalLatitude,
    decimalLongitude,
    COALESCE(coordinateUncertaintyInMeters, 1000)
  ) AS mgrsCode,
  speciesKey,
  species,
  family,
  -- Measurements
  COUNT(*) AS n,
  MIN(COALESCE(coordinateUncertaintyInMeters, 1000)) AS minCoordinateUncertaintyInMeters,
  -- Higher taxon measurement
  IF(ISNULL(family), NULL, SUM(COUNT(*)) OVER (PARTITION BY family)) AS familyCount
FROM
  birds_data.unstructured
GROUP BY
  year,
  mgrsCode,
  speciesKey,
  family,
  species
ORDER BY
  year ASC,
  mgrsCode ASC,
  speciesKey ASC;
```

The contributions of GBIF-mediated datasets:

```sql
SELECT
  -- Dimensions
  datasetKey,
  collectionCode,
  license,
  -- Measurement
  COUNT(*) AS n
FROM
  birds_data.unstructured
GROUP BY
  datasetKey,
  collectionCode,
  license;
```
